Alternative Methods for Retaining Explicit and Finding Implicit Sharing in Embedded DSLs

### 2023-02-14 paper/paper.tex
added more benchmarks to paper

The goal of detecting sharing is to generate a graph structure (like the AST in
the previous section) that share's common subexpressions. So we're essentially
going to generate a Directed Acyclic Graph (DAG) instead of an AST. For example

@@

Kiselyov's method for detecting implicit sharing in finally tagless style uses
hash-consing \cite{kiselyov:sharing}. Hash-consing is based on a bijection of
nodes and a set of identifiers, e.g.,
\begin{minted}{haskell}
data BiMap a -- abstract
lookup_key :: Ord a => a -> BiMap a -> Maybe Int
lookup_val :: Int -> BiMap a -> a
insert :: Ord a => a -> BiMap a -> (Int,BiMap a)
empty :: BiMap a
\end{minted}

@@

The representation for the finally tagless
instance is then a wrapper around a state monad that holds the DAG being constructed in its state and
returns the current (top) \mintinline{haskell}{NodeID}:
\begin{minted}{haskell}

newtype Graph a = Graph { unGraph :: State DAG NodeID }

instance Exp Graph where
  constant x = Graph (hashcons $ NConstant x)
  variable x = Graph (hashcons $ NVariable x)
  add e1 e2 = Graph (do
                     h1 <- unGraph e1
                     h2 <- unGraph e2
                     hashcons $ NAdd h1 h2)
\end{minted}
The trick to uncovering sharing is in the
\mintinline{haskell}{hashcons} function, which inserts a new node into the current DAG, but not
before checking if it is already there.
\begin{minted}{haskell}
hashcons :: Node -> State DAG NodeID
hashcons e = do
  DAG m <- get
  case lookup_key e m of
    Nothing -> let (k,m') = insert e m
               in put (DAG m') >> return k
    Just k -> return k
\end{minted}
The technique is essentially that of
hash-consing, popularized by its use in LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}. Other works have explored the use
of type-safe hash-consing in embedded DSLs, see \cite{filliatre:typesafeconsing}.

@@

However the method does not work for DSL functions
returning multiple outputs via tuples or container types like lists. Recall the definition
\begin{minted}{haskell}
novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}
The problem is that DAG generation requires splitting the state monad in two:
\begin{minted}{haskell}
instance Exp Graph where
  ...
  novel e1 e2 = let
     g1 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    hashcons $ Novel1 h1 h2)
     g2 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    hashcons $ Novel2 h1 h2)
     in (g1,g2)
\end{minted}
Each output it returns will now have to be individually evaluated, so a
chain of DSL functions that output 2 or more values will suffer from the same
exponential explosion of hashcons operations,
and trying to adapt the let construct above,
just creates another function with the same problem (multiple outputs).

@@

\section{Explicit Sharing Of ByteString ASTs}
We propose another solution to this issue, taking inspiration again from Kiselyov  
\cite{kiselyov:sharing}, by introducing an explicit construct for specifying
sharing. This time, the construct will substitute the current byte string for a
more compact label. For safety purposes, we need to keep track of a table of
these labels and their corresponding ASTs, to make sure we don't use the same label for different ASTs.
\begin{minted}{haskell}
data DAG = DAG { dagTrie :: Trie (Node,NodeID)
               , dagSubMap :: Map ByteString ByteString
               , dagMaxID :: Int
               } deriving Show

data Graph a = Graph { unGraph :: State DAG NodeID
                     , unStringAST :: ByteString
                     , unSubT :: Maybe ByteString }

class Substitute repr where
  subT :: ByteString -> repr a -> repr a
instance Substitute Graph where
  subT s' (Graph g s _) = Graph g s' (Just s)

@@

The \mintinline{haskell}{subT} operation substitutes the current byte string AST with a new one, and
we define a new operation \mintinline{haskell}{subTInsert} to check if
the label already exists in the cache map before inserting it.
\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST mSubt) =
      do DAG trie _ _ <- get
         let sAST' = case mSubt of
                       Just s -> s
                       Nothing -> sAST
         case Trie.lookup sAST' trie of
           Nothing -> sT -- error "missing ast"
           Just (node,nodeID) ->
              do subTInsert mSubt sAST (node,nodeID)
                 return nodeID
  in sequence $ map seqArg inps

subTInsert :: Maybe ByteString -> ByteString
           -> (Node, NodeID) -> State DAG ()
subTInsert Nothing  _ _  = return ()
subTInsert (Just s) sAST nodeID =
  do DAG trie subtMap _ <- get
     case Map.lookup sAST subtMap of
        Just sAST' -> if sAST == sAST'
                      then return ()
                      else error "tried to resubT"
        Nothing -> let cMap' = Map.insert sAST s subtMap
                       trie' = Trie.insert sAST nodeID trie
                  in modify (\dag -> dag { dagTrie = trie'
                                      , dagSubMap = cMap' })
\end{minted}
We need to make sure we don't attempt to insert the same substitution for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the state monad to prevent or modify the substitution.
In the toy example, compilation crashes, but we could catch an exception instead.
Either way it's up to the DSL
writer to ensure they don't reuse the same label.

@@

  \caption{Benchmarks of \mintinline{haskell}{addChains} example with full
    explicit sharing.}\label{tab1}
\end{table}

@@

>>>>

The goal of detecting sharing is to generate a graph structure (like the AST in
the previous section) that share's common subexpressions. So we're essentially
going to generate a Directed Acyclic Graph (DAG) instead of an AST. For example,
we can use the following DAG structure that explicitly references nodes by a
unique identifier

@@

Kiselyov's method for detecting implicit sharing in finally tagless style uses
hash-consing \cite{kiselyov:sharing}. Hash-consing is based on a bijection of
nodes and a set of identifiers, e.g.,

\begin{minted}{haskell}
data BiMap a -- abstract
lookup_key :: Ord a => a -> BiMap a -> Maybe Int
lookup_val :: Int -> BiMap a -> a
insert :: Ord a => a -> BiMap a -> (Int,BiMap a)
empty :: BiMap a
\end{minted}

@@

In order to generate such a data structure, we will need to keep track of the
current maximum identifier to keep them unique. The representation for the
finally tagless instance is then a wrapper around a state monad that holds the
DAG being constructed in its state and returns the current (top)
\mintinline{haskell}{NodeID}:

\begin{minted}{haskell}
newtype Graph a = Graph { unGraph :: State DAG NodeID }

instance Exp Graph where
  constant x = Graph (hashcons $ NConstant x)
  variable x = Graph (hashcons $ NVariable x)
  add e1 e2 = Graph (do
                     h1 <- unGraph e1
                     h2 <- unGraph e2
                     hashcons $ NAdd h1 h2)
\end{minted}

The trick to uncovering sharing is in the
\mintinline{haskell}{hashcons} function, which inserts a new node into the current DAG, but not
before checking if it is already there.
\begin{minted}{haskell}
hashcons :: Node -> State DAG NodeID
hashcons e = do
  DAG m <- get
  case lookup_key e m of
    Nothing -> let (k,m') = insert e m
               in put (DAG m') >> return k
    Just k -> return k
\end{minted}

The technique is essentially that of hash-consing, popularized by its use in
LISP compilers, but discovered by Ershov in 1958 \cite{ershov1958:consing}.
Other works have explored the use of type-safe hash-consing in embedded DSLs,
see \cite{filliatre:typesafeconsing}.

@@

However the method does not work for DSL functions returning multiple outputs
via tuples or container types like lists. If we're attempting to generate a more
specific graph structure like a hypergraph or bi-partite graph the need to do
this becomes more aparent. For example, in our own research we implemented a DSL
for an instruction set architecture that generates a bipartite graph of
instructions and the registers they act upon. Attempting to generate a graph
where an instruction outputs multiple registers will result in returning
multiple independent state monads.

We can more concisely illustrate this problem by adding the following
instruction to our DSL that attempts to return two separate nodes in the DAG at
once
\begin{minted}{haskell}
novel :: (repr Int,repr Int) -> (repr Int,repr Int)
\end{minted}

The issue is that DAG generation requires splitting the state monad in two:
\begin{minted}{haskell}
instance Exp Graph where
  ...
  novel e1 e2 = let
     g1 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    hashcons $ Novel1 h1 h2)
     g2 = Graph (do h1 <- unGraph e1
                    h2 <- unGraph e2
                    hashcons $ Novel2 h1 h2)
     in (g1,g2)
\end{minted}
Each output it returns will now have to be individually evaluated, so a chain of
DSL functions that output 2 or more values will suffer from the same exponential
explosion of hashcons operations, and trying to adapt the let construct above,
just creates another function with the same problem (multiple outputs).

@@

\section{Explicit Sharing Of ByteString ASTs}
We propose another solution to this issue, taking inspiration again from Kiselyov  
\cite{kiselyov:sharing}, by introducing an explicit construct for specifying
sharing. This time, the construct will substitute the current byte string for a
more compact label.

\begin{minted}{haskell}
class Substitute repr where
  subT :: ByteString -> repr a -> repr a
instance Substitute Graph where
  subT s' (Graph g s _) = Graph g s' (Just s)

@@

For safety purposes, we need to keep track of a table of these labels and their
corresponding ASTs, to make sure we don't use the same label for different ASTs.
\begin{minted}{haskell}
data DAG = DAG { dagTrie :: Trie (Node,NodeID)
               , dagSubMap :: Map ByteString ByteString
               , dagMaxID :: Int
               } deriving Show

data Graph a = Graph { unGraph :: State DAG NodeID
                     , unStringAST :: ByteString
                     , unSubT :: Maybe ByteString }
\end{minted}

When a substitution is made via \mintinline{haskell}{subT}, the
\mintinline{haskell}{unStringAST} field is replaced with the new label, and the
previous serialized AST is placed in \mintinline{haskell}{unSubT}. When a
\mintinline{haskell}{Graph} value is processed, the \mintinline{haskell}{unSubT}
field is checked for if it contains a label

\begin{minted}{haskell}
seqArgs :: [Graph a] -> State DAG [NodeID]
seqArgs inps =
  let
    seqArg (Graph sT sAST mSubt) =
      do DAG trie _ _ <- get
         let sAST' = case mSubt of
                       Just s -> s
                       Nothing -> sAST
         case Trie.lookup sAST' trie of
           Nothing -> sT -- error "missing ast"
           Just (node,nodeID) ->
              do subTInsert mSubt sAST (node,nodeID)
                 return nodeID
  in sequence $ map seqArg inps
\end{minted}

If the \mintinline{haskell}{unSubT} field contains a label, that means the
current \mintinline{haskell}{unStringAST} field is a substitution that needs to
be inserted into \mintinline{haskell}{dagSubMap}. The function
\mintinline{haskell}{subTInsert} handles this

\begin{minted}{haskell}
subTInsert :: Maybe ByteString -> ByteString
           -> (Node, NodeID) -> State DAG ()
subTInsert Nothing  _ _  = return ()
subTInsert (Just s) sAST nodeID =
  do DAG trie subtMap _ <- get
     case Map.lookup sAST subtMap of
        Just sAST' -> if sAST == sAST'
                      then return ()
                      else error "tried to resubT"
        Nothing -> let cMap' = Map.insert sAST s subtMap
                       trie' = Trie.insert sAST nodeID trie
                  in modify (\dag -> dag { dagTrie = trie'
                                      , dagSubMap = cMap' })
\end{minted}

We need to make sure we don't attempt to insert the same substitution for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the state monad to prevent or modify the substitution. In the toy example,
compilation crashes, but we could catch an exception instead. Either way it's up
to the DSL writer to ensure they don't reuse the same label as a substitution.

@@

  \vspace{1mm}
  \caption{Benchmarks of \mintinline{haskell}{addChains} example with full
    explicit sharing.}\label{tab1}
\end{table}

@@

For more interesting benchmarks, we need to consider a more sophisticated DSL.
We can define a DSL for an instruction set architecture like mentioned in
section \ref{limitexplicit}. It will have the following interface

\begin{minted}{haskell}
class ISA repr where
  -- | Load from memory into a GPR
  ldMR :: repr MR -> Int -> (repr GPR, repr MR)
  -- | Store a GPR into memory
  stdMR   :: repr MR -> Int -> repr GPR -> repr MR
  -- | Bitwise NAND of two 64-bit general-purpose registers (NNGRK)
  nandG   :: repr GPR -> repr GPR -> repr GPR
  -- | Bitwise NOR of two 64-bit general-purpose registers (NOGRK)
  norG    :: repr GPR -> repr GPR -> repr GPR
  -- | Bitwise NXOR of two 64-bit general-purpose registers (NXGRK)
  eqvG    :: repr GPR -> repr GPR -> repr GPR
  -- | Addition of two 64-bit general-purpose registers (AGRK)
  addG    :: repr GPR -> repr GPR -> repr GPR
  ...
\end{minted}

This language can be used to encode basic blocks of assembly code. These basic
blocks may return multiple outputs, preventing us from using explicit sharing
via let constructs. For example,
\begin{minted}{haskell}
add2 :: ISA repr => (repr GPR, repr GPR) -> (repr GPR, repr GPR)
add2 (a, b) =
    let
        a' = addG a b
        b' = addG a' b
    in (a', b')
\end{minted}

We used this language to implement approximations of vector cos and tan (i.e., a
loop body that computes over an input and output array). The loop body was
unrolled by a factor of four making for a sufficiently large resulting DAG.

We profiled cos using both Kiselyov's method (i.e., hash-consing) and are own
(i.e., trie-consing with some explicit sharing). It wasn't possible to use
Kiselyov's explicit sharing method where meaningful so our code was able to
achieve greater performance, with only a small increase in memory consumption
(see figure \ref{fig:hashvscons}).

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/cos_profile_hashcons.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/cos_profile_triecons.png}
  \caption{Comparison of cos using hashcons vs triecons}
  \label{fig:hashvscons}
\end{figure}

For tan, we were unable to generate a resulting code graph using Kiselyov's
method. Using explicit sharing via let constructs wasn't possible due to the
multiple outputs issue and without it the amount of computation due to redundant
hash-consing was simply too large to perform in a reasonable amount of time.

We were able to generate tan using our method both without and with our explicit
sharing, however without explicit sharing we consumed an unreasonable amount of
memory (see figure \ref{fig:explicitvsno}). It should be noted, the main source
of redundant computation in tan is the reuse of computed values of cos and sin.
By simply explicitly sharing just those values we achieved the significant
speedup shown in figure \ref{fig:explicitvsno}.

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/noexplicit.png}
  \hfill
  \includegraphics[width=0.45\textwidth]{figs/explicit.png}
  \caption{Comparison of cos using hashcons vs triecons}
  \label{fig:explicitvsno}
\end{figure}
